A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-11 tidyindex.qmd
a draft on introducing the general notation

The notation used for introducing the pipeline modules is as follows. Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspects: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$. When more than one variable is involved, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.

@@

**Input: ** $x(\mathbf{s};\mathbf{t})$: multivariate data with temporal coordinates

**Output: ** $x(\mathbf{s};\mathbf{t}^\prime)$: multivariate data with aggregated temporal axis

**Computation:** The construction of an index sometimes needs to consider information from neighbouring time periods.
The temporal processing is a general operator on the time dimension of the data in the form of

@@

**Input: ** Multivariate data with coordinates

**Output: ** Multivariate data with a new set of coordinates

**Computation:** Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of

@@

**Input: **

**Output: **

**Computation:** The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

@@

**Input: **

**Output: **

**Computation:** While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of

@@

**Input: **

**Output: **

**Computation:** Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

@@

*model fit? *

**Input: **

**Output: **

**Computation:** Distribution fit can be seen as the model fitting in its simplest term. It can be represented by

@@

**Input: **

**Output: **

**Computation:** This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
Also, the use of a normal scale needs to be interpreted with caution. 
@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

**Input: **

**Output: **

**Computation:** Benchmarking sets a constant value to allow the constructed index to be compared across time.
Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.

@@

**Input: **

**Output: **

**Computation:** In public communication, the index values are usually accompanied by a categorical grade. The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes. 
The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:

\begin{equation}
\begin{cases}
C_0 & c_1 \leq (s_i; t_j) < c_0 \\
C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
\cdots \\
C_z & c_z \leq x(s_i; t_j)
\end{cases}
\end{equation}

>>>>

Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal data,

\begin{equation}
\begin{aligned}
x^l(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^L(s;t)] \\
& l = 1, 2, \cdots L, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{aligned}
\end{equation}

where $x^l$ is the $l$th variable, $s$ is the spatial location, and $t$ is the temporal order. <!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.--> The spatial space, $D_s \subseteq \mathbb{R}^d$, is generic to include different spatial representations, including a 1D set of $n$ locations $\{D_s: s_1, s_2, \cdots, s_n\}$, a 2D set of longitude and latitude pairs, and $d$-D coordinate set. The higher dimension in the temporal space, $D_t \subseteq \mathbb{R}^m$, can be used to represent for a $m$-D deconstruction of time. For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$. In the case when the spatial dimension is fixed, multivariate temporal data can be simplified as $x^l_s(t)$, and similarly,  multivariate spatial data can be simplified as $x^l_t(s)$.

@@

`Input`: $x^l_s(t)$,

`Output`:  $x^l_s(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

`Computation`: The construction of an index sometimes needs to consider information from neighbouring time periods.
The temporal processing is a general operator on the time dimension of the data in the form of

@@

`Input`:  $x^l_t(s)$

`Output`: $x^l_t(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$

`Computation`: Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of

@@

`Input`: $x^l(s; t)$

`Output`: $y^l(s; t)$

`Computation`: The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

@@

`Input`: $x^l(s; t)$

`Output`: $x^l(s; t)$

`Computation`: While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of

@@

`Input`: $x^l(s; t)$

`Output`: $x^{l^\prime}(s; t)$, where $l^{\prime} = 1, 2, \cdots, L^{\prime}$

`Computation`: Dimension reduction summarises the multivariate information into univariate, which can be denoted as:

@@

<!-- *model fit? * -->

`Input`:  $x^l(s; t)$

`Output`:

`Computation`: Distribution fit can be seen as the model fitting in its simplest term. It can be represented by

@@

`Input`:  $x^l(s; t)$

`Output`:

`Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
Also, the use of a normal scale needs to be interpreted with caution. 
@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

`Input`:  $x^l(s; t)$

`Output`:

$D(x^l(s,t), a)$ where $D(., .)$ is a distance function between

`Computation`: Benchmarking sets a constant value to allow the constructed index to be compared across time.
Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.

@@

`Input`:  $x^l(s; t)$

`Output`: $C^l(s;t)$ where $\{C: c_0, c_1, \cdots, c_z\}$ is a discrete set

`Computation`: In public communication, the index values are usually accompanied by a categorical grade. The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes. 
The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:

\begin{equation}
\begin{cases}
c_0 & C_1 \leq x^l(s; t) < C_0 \\
c_1 & C_2 \leq x^l(s; t) < C_1 \\
c_2 & C_3 \leq x^l(s; t) < C_2 \\
\cdots \\
c_z & C_z \leq x^l(s; t)
\end{cases}
\end{equation}

### 2023-08-11 tidyindex.tex
a draft on introducing the general notation

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

The notation used for introducing the pipeline modules is as follows.
Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
spatial, temporal, and multivariate aspects: the spatial dimension
\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
defined in the 1D space:
\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
one variable is involved, the multivariate data can also be written as:
\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).

@@

\textbf{Input: } \(x(\mathbf{s};\mathbf{t})\): multivariate data with
temporal coordinates

\textbf{Output: } \(x(\mathbf{s};\mathbf{t}^\prime)\): multivariate data
with aggregated temporal axis

\textbf{Computation:} The construction of an index sometimes needs to
consider information from neighbouring time periods. The temporal
processing is a general operator on the time dimension of the data in
the form of

@@

\textbf{Input: } Multivariate data with coordinates

\textbf{Output: } Multivariate data with a new set of coordinates

\textbf{Computation:} Spatial processing may be needed when indexes are
not calculated independently on each collected location or when
variables collected from multiple sources need to be fused before
further processing. The process can be written as a general operation in
the form of

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} The purpose of variable transformation is to
create variables that fits assumptions for further computing. These
assumptions include a stable variance, normal distribution, or a certain
scale required by some algorithms down the pipeline. Variable
transformation is a general notion of a functional transformation on the
variable:

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} While scaling can be seen as a specific type of
variable transformation, it is separated into its own step to make the
step explicit in the pipeline. The key difference between the two steps
is that variable transformation typically changes the shape of the data
while scaling only changes the data scale and can usually be written in
the form of

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Dimension reduction summarises the multivariate
information into univariate, which can be denoted as:

@@

\emph{model fit? }

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Distribution fit can be seen as the model fitting
in its simplest term. It can be represented by

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} This step maps the univariate series into a
different scale, typically for ease of comparison across regions. For
example, a normal scale, {[}0, 1{]}, or {[}0, 100{]} may be favored for
reporting certain indexes. In drought indexes, i.e.~SPI or SPEI, the
quantiles from the fitted distribution are converted into the normal
scale via the normal reverse CDF function: \(\Phi^{-1}(.)\). Normalising
is usually used at the end of the pipeline and its main difference from
the scaling step is that here the change of scale also changes the
distribution of the variable. While being commonly used, this step can
get criticism from analysts for forcing the data into the decided scale,
which can be either unnecessary or inaccurately exaggerate or downplay
the outliers. Also, the use of a normal scale needs to be interpreted
with caution. Figure~\ref{fig-normalising} illustrates the normal
density not being directly proportional to its probability of
occurrence. This is concerning, especially at the extreme values, since
a small difference in the tail density can have magnitudes of difference
in its probability of occurrence.

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} Benchmarking sets a constant value to allow the
constructed index to be compared across time. Here we denote it with
\(u[x(s_i, t_j)]\) where \(u\) is a scalar of interest in the index
constructed. A benchmark value could be a constant or a function of the
data, i.e.~mean.

@@

\textbf{Input: }

\textbf{Output: }

\textbf{Computation:} In public communication, the index values are
usually accompanied by a categorical grade. The categorised grades are
an ordered set of descriptive words or colors to communicate the
severity or guide the comprehension of the indexes. The mapping from
continuous index values to the discrete grades is called simplification
in the pipeline and it can be written as a piece-wise function:

\begin{equation}
\begin{cases}
C_0 & c_1 \leq (s_i; t_j) < c_0 \\
C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
\cdots \\
C_z & c_z \leq x(s_i; t_j)
\end{cases}
\end{equation}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Now, we introduce the notation used for describing pipeline modules.
Consider multivariate spatio-temporal data,

\begin{equation}
\begin{aligned}
x^l(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^L(s;t)] \\
& l = 1, 2, \cdots L, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{aligned}
\end{equation}

where \(x^l\) is the \(l\)th variable, \(s\) is the spatial location,
and \(t\) is the temporal order. The spatial space,
\(D_s \subseteq \mathbb{R}^d\), is generic to include different spatial
representations, including a 1D set of \(n\) locations
\(\{D_s: s_1, s_2, \cdots, s_n\}\), a 2D set of longitude and latitude
pairs, and \(d\)-D coordinate set. The higher dimension in the temporal
space, \(D_t \subseteq \mathbb{R}^m\), can be used to represent for a
\(m\)-D deconstruction of time. For example, a year-month combination,
\((Y, M)^\prime\), can be used to record monthly precipitation series
from January 1990 to December 2020, where \(Y = 1990, \cdots, 2020\) and
\(M = 1, \cdots, 12\). In the case when the spatial dimension is fixed,
multivariate temporal data can be simplified as \(x^l_s(t)\), and
similarly, multivariate spatial data can be simplified as \(x^l_t(s)\).

@@

\texttt{Input}: \(x^l_s(t)\),

\texttt{Output}: \(x^l_s(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

\texttt{Computation}: The construction of an index sometimes needs to
consider information from neighbouring time periods. The temporal
processing is a general operator on the time dimension of the data in
the form of

@@

\texttt{Input}: \(x^l_t(s)\)

\texttt{Output}: \(x^l_t(s^\prime)\) where
\(s^\prime \in D_{s^{\prime}}\)

\texttt{Computation}: Spatial processing may be needed when indexes are
not calculated independently on each collected location or when
variables collected from multiple sources need to be fused before
further processing. The process can be written as a general operation in
the form of

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}: \(y^l(s; t)\)

\texttt{Computation}: The purpose of variable transformation is to
create variables that fits assumptions for further computing. These
assumptions include a stable variance, normal distribution, or a certain
scale required by some algorithms down the pipeline. Variable
transformation is a general notion of a functional transformation on the
variable:

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}: \(x^l(s; t)\)

\texttt{Computation}: While scaling can be seen as a specific type of
variable transformation, it is separated into its own step to make the
step explicit in the pipeline. The key difference between the two steps
is that variable transformation typically changes the shape of the data
while scaling only changes the data scale and can usually be written in
the form of

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}: \(x^{l^\prime}(s; t)\), where
\(l^{\prime} = 1, 2, \cdots, L^{\prime}\)

\texttt{Computation}: Dimension reduction summarises the multivariate
information into univariate, which can be denoted as:

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}:

\texttt{Computation}: Distribution fit can be seen as the model fitting
in its simplest term. It can be represented by

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}:

\texttt{Computation}: This step maps the univariate series into a
different scale, typically for ease of comparison across regions. For
example, a normal scale, {[}0, 1{]}, or {[}0, 100{]} may be favored for
reporting certain indexes. In drought indexes, i.e.~SPI or SPEI, the
quantiles from the fitted distribution are converted into the normal
scale via the normal reverse CDF function: \(\Phi^{-1}(.)\). Normalising
is usually used at the end of the pipeline and its main difference from
the scaling step is that here the change of scale also changes the
distribution of the variable. While being commonly used, this step can
get criticism from analysts for forcing the data into the decided scale,
which can be either unnecessary or inaccurately exaggerate or downplay
the outliers. Also, the use of a normal scale needs to be interpreted
with caution. Figure~\ref{fig-normalising} illustrates the normal
density not being directly proportional to its probability of
occurrence. This is concerning, especially at the extreme values, since
a small difference in the tail density can have magnitudes of difference
in its probability of occurrence.

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}:

\(D(x^l(s,t), a)\) where \(D(., .)\) is a distance function between

\texttt{Computation}: Benchmarking sets a constant value to allow the
constructed index to be compared across time. Here we denote it with
\(u[x(s_i, t_j)]\) where \(u\) is a scalar of interest in the index
constructed. A benchmark value could be a constant or a function of the
data, i.e.~mean.

@@

\texttt{Input}: \(x^l(s; t)\)

\texttt{Output}: \(C^l(s;t)\) where \(\{C: c_0, c_1, \cdots, c_z\}\) is
a discrete set

\texttt{Computation}: In public communication, the index values are
usually accompanied by a categorical grade. The categorised grades are
an ordered set of descriptive words or colors to communicate the
severity or guide the comprehension of the indexes. The mapping from
continuous index values to the discrete grades is called simplification
in the pipeline and it can be written as a piece-wise function:

\begin{equation}
\begin{cases}
c_0 & C_1 \leq x^l(s; t) < C_0 \\
c_1 & C_2 \leq x^l(s; t) < C_1 \\
c_2 & C_3 \leq x^l(s; t) < C_2 \\
\cdots \\
c_z & C_z \leq x^l(s; t)
\end{cases}
\end{equation}
